quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space

### 2025-12-19 Archive/paper-quollr.Rmd
Updated spellings and try to make submit .zip file

```{r model-limb, fig.cap="Compare the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-hbe}a) and the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-hbe}f) by HBE plot (Figure \\ref{fig:limb-hbe}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.alt = "The figure shows a 2-D scatter plot of a tSNE layout for the Limb muscle dataset (n=1067) with seven clusters in 10 dimensions. Each point represents one observation, with point colour indicating the cluster label reported in the original paper. The horizontal and vertical axes represent the two tSNE coordinates, both on roughly symmetric numeric scales around zero. Points form several visually distinct clusters, with some local variation visible within clusters. A model overlay with a binwidth of 0.06 shows the local density structure, providing a background grid that highlights differences between clusters without fragmenting the plot.", fig.pos="!ht", fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

>>>>

```{r model-limb, fig.cap="Representative views of two selected NLDR layouts for the Limb muscle dataset ($n=1067$), shown row-wise. The top row (a1–a3) corresponds to the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-hbe}a), and the bottom row (f1–f3) corresponds to the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-hbe}f) selected using the HBE plot. In each row, the left panel (a1, f1) shows the NLDR embedding with points coloured by muscle group and overlaid with triangulated hexagon centroids. The middle (a2, f2) and right (a3, f3) panels show two different $2\\text{-}D$ projections of the fitted model and data in the original $10\\text{-}D$ space, with the same triangular mesh displayed. Together, these panels summarize how the low-dimensional layouts relate to the underlying high-dimensional structure across different viewing directions.", fig.alt = "The figure consists of six panels arranged in two rows and three columns, showing representative views of two NLDR layouts for the Limb muscle dataset (n=1067). In the top row (a1–a3), the left panel (a1) shows the published 2-D NLDR embedding with points coloured by muscle group and overlaid with triangulated hexagon centroids. The middle (a2) and right (a3) panels show two different 2-D projections of the corresponding 10-D data and fitted model, with points plotted as individual markers and the same triangular mesh connecting centroids. In the bottom row (f1–f3), the left panel (f1) shows the 2-D NLDR layout selected using the HBE plot, again with coloured points and triangulated centroids, while panels (f2) and (f3) show two different 2-D projections of the associated 10-D structure with the same mesh displayed. Across all panels, axes represent continuous numeric coordinates, colour encodes muscle group, and the triangular mesh is used consistently to show correspondence between the low-dimensional layouts and the high-dimensional projections.", fig.pos="!ht", fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

### 2025-12-19 Archive/paper-quollr.tex
Updated spellings and try to make submit .zip file

\begin{figure}[H]
\includegraphics[width=1\linewidth,alt={Flowchart diagram illustrating the quollr workflow. It shows inputs of NLDR and high-dimensional data progressing through preprocessing, hexagonal binning, centroid computation, and triangulation to form a 2-D mesh. The mesh is lifted into the high-dimensional space for prediction and error analysis, with interactive tools linking the high- and low-dimensional representations.}]{figures/quollr_workflow} \caption{Overview of the `quollr` workflow and software architecture. The process begins with NLDR and $p\text{-}D$ data inputs, followed by data preprocessing and hexagonal binning. Centroids are computed and triangulated to form the $2\text{-}D$ mesh, which is then lifted into the $p\text{-}D$ space. Predictions and error computations are performed on new data, while interactive functions enable dynamic linking between the $p\text{-}D$ and $2\text{-}D$ representations.}\label{fig:workflow}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Four-panel figure showing the main steps in building the model on a UMAP layout using S-curve data: (a) hexagonal binning of points, (b) calculation of bin centroids, (c) triangulation of centroids to form a mesh, and (d) lifting the triangulated model into the high-dimensional space.}]{paper-quollr_files/figure-latex/algo-steps-1}

@@

{\centering \includegraphics[width=1\linewidth,alt={Four-panel figure showing outputs from geom_hexgrid and geom_trimesh functions: (a) the full hexagonal grid, (b) the subset of the grid that overlaps with the data, (c) a complete triangulated grid based on centroids, and (d) the triangulated centroid grid corresponding to the data region.}]{paper-quollr_files/figure-latex/geom-outputs-pdf-1}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Four-panel figure showing 2-D projections of the lifted high-dimensional wireframe model from the S-curve UMAP layout. Each panel displays the black wireframe model overlaid on blue S-curve data points, illustrating how the model captures the curved structure and revealing regions with sparse or no data.}]{paper-quollr_files/figure-latex/scurve-highd-model-pdf-1} \caption{$2\text{-}D$ projections of the lifted high-dimensional wireframe model from the \texttt{scurve} UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on \texttt{scurve} data (blue) in different projections. These views illustrate how the lifted wireframe model captures the structure of the \texttt{scurve} data. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.}\label{fig:scurve-highd-model-pdf}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=0.5\linewidth,alt={Screenshots from the detourr interface showing the lifted high-dimensional wireframe model derived from the S-curve UMAP layout. The images highlight how the wireframe represents the data structure and expose regions with sparse or missing data.}]{figures/model_proj1_detourr} \includegraphics[width=0.5\linewidth,alt={Screenshots from the detourr interface showing the lifted high-dimensional wireframe model derived from the S-curve UMAP layout. The images highlight how the wireframe represents the data structure and expose regions with sparse or missing data.}]{figures/model_proj2_detourr} \caption{Screenshots of the lifted high-dimensional wireframe model from the \texttt{scurve} UMAP layout using \texttt{detourr}. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.}\label{fig:scurve-highd-model-detourrpdf}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Four-panel figure with linked views showing correspondence between the UMAP layout and the S-curve structure in seven dimensions. Panels (a1, b1) show selected regions in the UMAP layout, and panels (a2, b2) display the corresponding points in the high-dimensional tour view. Purple points indicate selected subsets: the lower bridge of the S-curve in (a1, a2) and the upper arm region in (b1, b2). The figure illustrates how continuous structures in the 7-D S-curve appear as separate clusters in the 2-D UMAP layout, revealing distortion effects.}]{paper-quollr_files/figure-latex/scurve-nldrlink-pdf-1} \caption{Exploring the correspondence between UMAP layout and \texttt{scurve} structure in $7\text{-}D$. Two sets of plots are interactively linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the \texttt{scurve} is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the \texttt{scurve} is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:scurve-nldrlink-pdf}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,alt={Six-panel figure showing linked views of residuals and UMAP layouts for the 7-D S-curve model. Panels (a1–a3) highlight high-residual points in purple, corresponding to sparse bridging regions in the UMAP layout and less dense ends of the high-dimensional S-curve. Panels (b1–b3) highlight low-residual points, mapping to denser regions in the UMAP and thicker bands in the high-dimensional projection. The figure illustrates how residuals reveal distortions in UMAP embeddings.}]{paper-quollr_files/figure-latex/scurve-linkerror-pdf-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ \texttt{scurve} model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the \texttt{scurve} in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the \texttt{scurve} in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:scurve-linkerror-pdf}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Comparison of published $2\text{-}D$ layout (b) and the HBE-selected $2\text{-}D$ layout (f) for the limb muscle data (n = 1067) with seven differently shaped clusters in $10\text{-}D$, using tSNE, UMAP, PHATE, TriMAP, and PaCMAP with various hyper-parameters.}]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2\text{-}D$ layout (Figure \ref{fig:limb-hbe}a) and the $2\text{-}D$ layout selected (Figure \ref{fig:limb-hbe}f) by HBE plot (Figure \ref{fig:limb-hbe}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

>>>>

\begin{figure}[H]
\includegraphics[width=1\linewidth,alt={The figure is a schematic diagram illustrating the quollr workflow and software architecture. Rectangular boxes represent processing steps and data objects, including NLDR and p-D data inputs, data preprocessing, hexagonal binning, centroid computation, triangulation, and construction of a 2-D mesh. Arrows indicate the flow of information between steps, from the input data through the 2-D mesh, and then from the mesh being lifted into the p-D space. Additional boxes show prediction and error computation for new data, as well as interactive components that link the p-D and 2-D representations. The diagram is laid out sequentially, showing how data move through the system from input to interactive output.}]{figures/quollr_workflow} \caption{Overview of the `quollr` workflow and software architecture. The process begins with NLDR and $p\text{-}D$ data inputs, followed by data preprocessing and hexagonal binning. Centroids are computed and triangulated to form the $2\text{-}D$ mesh, which is then lifted into the $p\text{-}D$ space. Predictions and error computations are performed on new data, while interactive functions enable dynamic linking between the $p\text{-}D$ and $2\text{-}D$ representations.}\label{fig:workflow}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={The figure is a multi-panel diagram illustrating four steps in constructing a model on a UMAP layout using the scurve data. Panel (a) shows a 2-D UMAP scatter plot with points grouped into hexagonal bins covering the layout. Panel (b) shows the same layout with a centroid point displayed for each hexagon bin. Panel (c) shows the centroids connected by straight line segments to form a triangulated mesh across the 2-D layout. Panel (d) shows this 2-D triangulated mesh lifted into the original high-dimensional space, where the mesh appears as a curved surface. The horizontal and vertical axes in the 2-D panels represent the two UMAP coordinates with bounded numeric ranges, while the lifted view represents coordinates in the higher-dimensional space without explicit axis scales.}]{paper-quollr_files/figure-latex/algo-steps-1}

@@

{\centering \includegraphics[width=1\linewidth,alt={The figure consists of four 2-D panels illustrating hexagonal grids and triangulated meshes. Panel (a) shows a complete hexagonal grid covering a rectangular 2-D region, with all hexagons drawn regardless of whether data are present. Panel (b) shows a hexagonal grid restricted to bins that contain data points, with only those hexagons displayed and gaps where bins are empty. Panel (c) shows a full triangulated mesh constructed from centroids across the entire grid region, with centroids connected by straight line segments to form triangles. Panel (d) shows a triangulated mesh constructed only from centroids corresponding to non-empty bins, aligned with the data layout and leaving open areas where no centroids are present. In all panels, the horizontal and vertical axes represent continuous 2-D coordinates with bounded numeric ranges, and the grids and meshes are drawn with uniform line styles and colours.}]{paper-quollr_files/figure-latex/geom-outputs-pdf-1}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure shows four panels of 2-D scatter plots (a1–a4), each displaying a different projection of the lifted high-dimensional wireframe model derived from the scurve UMAP layout. In each panel, the horizontal and vertical axes represent two projected coordinates with continuous numeric ranges. The wireframe model is drawn in black as a wireframe of connected line segments, and the scurve data points are overlaid in blue as individual markers. The relative positions and orientations of the wireframe and data points vary across panels as different projections are shown. In some regions, the wireframe extends into areas with few or no data points, leaving visible gaps between the model and the data. All panels have similar aspect ratios and use consistent colour and symbol encodings to distinguish the model from the data.}]{paper-quollr_files/figure-latex/scurve-highd-model-pdf-1} \caption{$2\text{-}D$ projections of the lifted high-dimensional wireframe model from the \texttt{scurve} UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on \texttt{scurve} data (blue) in different projections. These views illustrate how the lifted wireframe model captures the structure of the \texttt{scurve} data. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.}\label{fig:scurve-highd-model-pdf}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=0.5\linewidth,alt={The figure shows several screenshots of 2-D projections of the lifted high-dimensional wireframe model derived from the scurve UMAP layout using detourr. In each screenshot, the horizontal and vertical axes represent projected coordinates with continuous numeric ranges. The wireframe model is displayed as a set of connected line segments, and the scurve data points are shown as individual markers overlaid on the wireframe. The orientation and shape of the wireframe and point cloud differ across screenshots as different projections are shown. In some views, parts of the wireframe extend into areas with few or no data points, leaving visible gaps between the model and the data. All screenshots use consistent colour and line styles and have similar aspect ratios.}]{figures/model_proj1_detourr} \includegraphics[width=0.5\linewidth,alt={The figure shows several screenshots of 2-D projections of the lifted high-dimensional wireframe model derived from the scurve UMAP layout using detourr. In each screenshot, the horizontal and vertical axes represent projected coordinates with continuous numeric ranges. The wireframe model is displayed as a set of connected line segments, and the scurve data points are shown as individual markers overlaid on the wireframe. The orientation and shape of the wireframe and point cloud differ across screenshots as different projections are shown. In some views, parts of the wireframe extend into areas with few or no data points, leaving visible gaps between the model and the data. All screenshots use consistent colour and line styles and have similar aspect ratios.}]{figures/model_proj2_detourr} \caption{Screenshots of the lifted high-dimensional wireframe model from the \texttt{scurve} UMAP layout using \texttt{detourr}. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.}\label{fig:scurve-highd-model-detourrpdf}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure consists of four linked panels arranged in two rows and two columns. Panels (a1) and (b1) show 2-D scatter plots of a UMAP layout, with the horizontal and vertical axes representing the two UMAP coordinates on continuous numeric scales. Panels (a2) and (b2) show corresponding 2-D projections from a tour of the 7-D data and model, where each axis represents a linear combination of the original seven dimensions. Data points are plotted as individual markers, and selected subsets are highlighted in purple while unselected points appear in a contrasting colour. In the top row, a set of points highlighted in the lower region of the UMAP layout in (a1) is linked to highlighted points in the tour view in (a2). In the bottom row, a different highlighted region in the UMAP layout in (b1) is linked to a different set of highlighted points in the tour view in (b2). The linked selections update consistently across panels, showing how the same subsets of observations appear in the UMAP layout and in different projections of the high-dimensional space.}]{paper-quollr_files/figure-latex/scurve-nldrlink-pdf-1} \caption{Exploring the correspondence between UMAP layout and \texttt{scurve} structure in $7\text{-}D$. Two sets of plots are interactively linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the \texttt{scurve} is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the \texttt{scurve} is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:scurve-nldrlink-pdf}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,alt={The figure consists of six panels arranged in two rows and three columns, showing residuals and their correspondence between UMAP and high-dimensional projections of the scurve data. In each row, the left panel (a1, b1) is a univariate plot of model residuals, with the horizontal axis representing residual values and the vertical axis representing frequency or density; points selected for highlighting appear in purple. The middle panel (a2, b2) shows a 2-D UMAP layout, with horizontal and vertical axes representing the two UMAP coordinates; the selected points from the left panel are highlighted in purple, while unselected points appear in a contrasting colour. The right panel (a3, b3) shows a 2-D projection of the 7-D data and model using a tour view, where axes represent linear combinations of the original dimensions; highlighted points correspond to those selected in the residuals and UMAP panels, while unselected points are shown in a contrasting colour. Across the panels, the purple highlights differ between the top and bottom rows, showing different subsets of points. All panels maintain roughly square aspect ratios and use consistent visual encodings for selection and non-selection.}]{paper-quollr_files/figure-latex/scurve-linkerror-pdf-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ \texttt{scurve} model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the \texttt{scurve} in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the \texttt{scurve} in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:scurve-linkerror-pdf}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure consists of six panels arranged in two rows and three columns, showing representative views of two NLDR layouts for the Limb muscle dataset (n=1067). In the top row (a1–a3), the left panel (a1) shows the published 2-D NLDR embedding with points coloured by muscle group and overlaid with triangulated hexagon centroids. The middle (a2) and right (a3) panels show two different 2-D projections of the corresponding 10-D data and fitted model, with points plotted as individual markers and the same triangular mesh connecting centroids. In the bottom row (f1–f3), the left panel (f1) shows the 2-D NLDR layout selected using the HBE plot, again with coloured points and triangulated centroids, while panels (f2) and (f3) show two different 2-D projections of the associated 10-D structure with the same mesh displayed. Across all panels, axes represent continuous numeric coordinates, colour encodes muscle group, and the triangular mesh is used consistently to show correspondence between the low-dimensional layouts and the high-dimensional projections.}]{paper-quollr_files/figure-latex/model-limb-1} \caption{Representative views of two selected NLDR layouts for the Limb muscle dataset ($n=1067$), shown row-wise. The top row (a1–a3) corresponds to the published $2\text{-}D$ layout (Figure \ref{fig:limb-hbe}a), and the bottom row (f1–f3) corresponds to the $2\text{-}D$ layout selected (Figure \ref{fig:limb-hbe}f) selected using the HBE plot. In each row, the left panel (a1, f1) shows the NLDR embedding with points coloured by muscle group and overlaid with triangulated hexagon centroids. The middle (a2, f2) and right (a3, f3) panels show two different $2\text{-}D$ projections of the fitted model and data in the original $10\text{-}D$ space, with the same triangular mesh displayed. Together, these panels summarize how the low-dimensional layouts relate to the underlying high-dimensional structure across different viewing directions.}\label{fig:model-limb}
\end{figure}
